### 2021-07-23 paper/localpsf.tex
H-matrix and discretization section

\section{Introduction}

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $d=1$, $2$, or $3$, and let $\mathcal{A}:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the form
\begin{equation}
\label{eq:kernel_representation}
(\mathcal{A}u)(v) := \int_\Omega \int_\Omega v(y) \Phi(y,x) u(x) dx dy,
\end{equation}
with integral kernel $\Phi:\Omega \times \Omega \rightarrow \mathbb{R}$. Here $\mathcal{A}u \in L^2(\Omega)'$ is the linear functional that results from applying $\mathcal{A}$ to $u\in L^2(\Omega)$, and $\left(\mathcal{A}u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We focus on operators of this form which have the following properties:
\begin{description}
\item[Matrix-free:] One cannot easily access kernel entries $\Phi(y,x)$. Access to $\mathcal{A}$ is available only through application of $\mathcal{A}$ and $\mathcal{A}^T$ to arbitrary functions (i.e., operator actions). That is, evaluation of the maps
\begin{equation*}
	u \mapsto\mathcal{A}u \quad \text{and} \quad v \mapsto\mathcal{A}^Tv.
\end{equation*} 
\item[Non-negative kernel:] For all $(y,x) \in \Omega \times \Omega$, we have
\begin{equation*}
	\Phi(y,x) \ge 0.
\end{equation*}
\item[Local support:] Let $\phi_x$ be the following \emph{impulse response}:
\begin{equation}
\label{eq:impulse_response_defn}
\phi_x(y) := \Phi(y, x).
\end{equation} 
We say that $\mathcal{A}$ has local support if, for all $x\in \Omega$, the support of $\phi_x$ is contained (or approximately contained) in a neighborhood of $x$. The smaller these supports are, the better our algorithm will perform.
\item[Local translation invariance:] If $h$ is not too large, then
\begin{equation}
	\label{eq:local_translation_invariance}
	\Phi(y+h, x+h) \approx \Phi(y,x).
\end{equation}
\end{description}

@@

\begin{thm}[Optimal interpolation]
	\label{thm:smoothest_interpolant}
Let $y \in \mathbb{R}^d$, and let $\gamma_y$ be the function
\begin{equation}
\label{eq:u_defn_in_thm}
\gamma_y(x) := \sum_{i=1}^r \varphi_i(y)w_i(x) = \widetilde{\Phi}\left(y+x,x\right),
\end{equation}
where the weighting functions $w_i$ are the solutions to optimization problem \eqref{eq:wi_optimization_problem}. Then $\gamma_y$ solves the optimization problem
\begin{equation}
\label{eq:uy_optimization_problem}
\begin{aligned}
\min_{\gamma_y \in H^2_N(\Omega)} &\quad \frac{1}{2}\norm{\Delta \gamma_y}_{L^2(\Omega)}^2 \\
\text{such that} &\quad \gamma_y(x_i) = \varphi_i(y), \quad i=1,\dots,q.
\end{aligned}
\end{equation}
\end{thm}

@@

We now describe how to efficiently evaluate $\widetilde{\Phi}(y,x)$ at an arbitrary point $(x,y) \in \Omega \times \Omega$. From \eqref{eq:kernel_entries}, we have
\begin{equation}
\label{eq:eval_y_x}
\widetilde{\Phi}(y,x) = \sum_{i=1}^r w_i(x) \varphi_{x_i}(y-x),
\end{equation}
and from \eqref{eq:varphi_eval}, we have
\begin{equation}
\label{eq:kernel_entry_formua}
\varphi_{x_i}(y-x) = \begin{cases}
\eta^{b(i)}(y - x + x_i), & y - x + x_i \in E_{x_i} \\
0, & \text{otherwise},
\end{cases}
\end{equation}
where $b(i)$ denotes the batch index such that $x_i \in S^{b(i)}$. To compute $z=\widetilde{\Phi}(y,x)$, we start with $z=0$, then loop through all batches $b$, and all sample points in each batch. For each sample point, we add one term in the sum in \eqref{eq:eval_y_x} to $z$, using \eqref{eq:kernel_entry_formua} as needed to evaluate $\varphi_{x_i}$. This is shown in Algorithm \ref{alg:eval_y_x}.

@@

\subsection{Conversion to hierarchical matrix format}
\label{sec:H_matrix_conversion}

We convert $\widetilde{\mathbf{A}}$ (the discretized version of $\widetilde{\mathcal{A}}$) to hierarchical matrix (H-matrix) format. Once in H-matrix format, useful matrix operations such as matrix factorization and inversion may be performed quickly using fast H-matrix methods. We use H1 matrices in our numerical results, but any of the other H-matrix formats (such as H2, HODLR, HSS, HBS, and others \cite{HMATRIX}) could be used instead. The basic ideas behind H-matrices are as follows:
\begin{itemize} 
	\item Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank.
	\item Recursive algorithms can take advantage of this low rank block structure to perform matrix arithmetic fast. Conventional algorithms for matrix inversion, matrix factorization, matrix-vector products, matrix-matrix products, and matrix addition require either $O(N^2)$ or $O(N^3)$ time and memory, while the aforementioned recursive algorithms can perform these matrix operations in $O(N \log^a N)$ time and memory for hierarchical matrices. Here $N$ is the number of degrees of freedom for the discretized version of the problem, and $a=0, 1, 2$, or $3$ depending on the operation and type of H-matrix.
\end{itemize}
For more details on H-matrices, we recommend \cite{HMATRIXGOOD}.

We convert $\widetilde{\mathbf{A}}$ into H1 matrix format using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package \cite{HLIBPRO}. Briefly, this method uses geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix to construct a recursive block partitioning of $\widetilde{\mathbf{A}}$, such that all blocks at the finest level of the partitioning are either low rank, or are small. Then, the method forms low rank approximations of the blocks that are expected to be low rank by sampling specially chosen rows and columns of those blocks. The small high rank blocks are formed directly by evaluating their entries. The primary cost of constructing the approximation is the evaluation of $O(N \log N)$ entries of the kernel $\widetilde{A}$, which we perform using Algorithm \ref{alg:eval_y_x}. We describe the H-matrix construction in detail in Appendix \ref{app:h_matrix}.

@@

In many problems of practical interest (e.g., Hessian approximation) $\mathcal{A}$ is symmetric positive semi-definite. However, $\widetilde{\mathbf{A}}$ is generally non-symmetric and indefinite because of errors in the product-convolution approximation. 
This is undesirable. Symmetry and positive semi-definiteness are important properties which should be preserved if possible. Also, lacking these properties may prevent one from using highly effective algorithms, such as the conjugate gradient method, to perform further useful operations involving $\widetilde{\mathbf{A}}$.
In this section we describe a method for modifying $\widetilde{\mathbf{A}}$ to make it symmetric positive semi-definite.

First, we use fast H-matrix addition to symmetrize $\widetilde{\mathbf{A}}$:
$$\widetilde{\mathbf{A}}_\text{sym} := \frac{1}{2}\left(\widetilde{\mathbf{A}} + \widetilde{\mathbf{A}}^T\right).$$
Then we modify $\widetilde{\mathbf{A}}_\text{sym}$ to make it positive semi-definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\widetilde{\mathbf{A}}_\text{sym}^+ := c_0 \mathbf{I} + c_1 \widetilde{\mathbf{A}}_\text{sym} + c_2 \left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\widetilde{\mathbf{A}}_\text{sym}$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ approximately equal the corresponding moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, while the negative and small positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ are modified so that $\widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. Later in this section we will explain how $c_0$, $c_1$, $c_2$, and $\mu$ are chosen and how our choice of these scalars affects the spectral properties of $\widetilde{\mathbf{A}}_\text{sym}^+$. In Figure FIG we compare the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ to that of $\widetilde{\mathbf{A}}_\text{sym}^+$ for different choices of these scalars. Forming $\widetilde{\mathbf{A}}_\text{sym}^+$ requires computing an H-matrix inverse, $\left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should usually be avoided, here it is computationally acceptable because $\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}$ is an H-matrix. Inversion of H-matrices is a routine procedure which only requires $O(N k^2 \log(N)^2)$ work, where $k$ is the H-matrix rank \cite{Hmatrixinverse}.

Approximation \eqref{eq:rational_matrix_function} may be written as $\widetilde{\mathbf{A}}_\text{sym}^+ = f(\widetilde{\mathbf{A}}_\text{sym})$, where
\begin{equation}
\label{eq:rational_function_scalar}
f(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\mu}.
\end{equation}
We choose $c_0, c_1, c_2$, and $\mu$, to be the solution to the following system of equations:
\begin{subequations}
	\label{eq:four_eq_system}
\begin{align}
f(0) &= 0 \label{eq:desired_property_1} \\
f'(0) &= 0 \label{eq:desired_property_2} \\
f(\lambda_\text{min}) &= |\lambda_\text{min}| \label{eq:desired_property_3} \\
f(\lambda_\text{max}) &= \lambda_\text{max}, \label{eq:desired_property_4}
\end{align}
\end{subequations}
where $\lambda_\text{min}$ and $\lambda_\text{max}$ are the smallest and largest eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, respectively. We also require $\mu > |\lambda_\text{min}|$, so that the pole of $f$ is not in $[\lambda_\text{min},\lambda_\text{max}]$. Note that $\lambda_\text{min} < 0$ since $\widetilde{\mathbf{A}}_\text{sym}$ is indefinite. 
We estimate $\lambda_\text{min}$ and $\lambda_\text{max}$ using the implicitly restarted Lanczos method \cite{lanczos,scipy}. 
%This is fast because computing matrix-vector products with an H-matrix is fast, and because the eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ do not cluster near $\lambda_\text{min}$ or $\lambda_\text{max}$ (though the eigenvalues may cluster at zero).

Together, equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, \eqref{eq:desired_property_3},  and \eqref{eq:desired_property_4}, and the condition $\mu > |\lambda_\text{min}|$, imply that $f(0)=0$ is the unique minimum of $f$ on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that $f$ is non-negative on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that  $f(\widetilde{\mathbf{A}}_\text{sym}) = \widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. Furthermore, \eqref{eq:desired_property_1} implies that the null space of $\widetilde{\mathbf{A}}_\text{sym}^+$ contains the null-space of $\widetilde{\mathbf{A}}_\text{sym}$. This is important when the operator $\mathcal{A}$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. Equation \eqref{eq:desired_property_4} forces the moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ to be close to the corresponding moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$. I.e., $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. Equation \eqref{eq:desired_property_3} ensures that the negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $f$. There is a tradeoff: if we choose a larger value for $f(\lambda_\text{min})$, then the positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ will be closer to the corresponding positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ (good), but the erroneous negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ will be amplified in magnitude (bad). This is shown in Figure FIG. We find that $f(\lambda_\text{min})=|\lambda_\text{min}|$ is an appropriate balance between these competing interests.

>>>>

\section{Introduction}
\label{sec:intro}

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $d=1$, $2$, or $3$, and let $\mathcal{A}:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the form
\begin{equation}
\label{eq:kernel_representation}
(\mathcal{A}u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy,
\end{equation}
with integral kernel $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$. Here $\mathcal{A}u \in L^2(\Omega)'$ is the linear functional that results from applying $\mathcal{A}$ to $u\in L^2(\Omega)$, and $\left(\mathcal{A}u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We focus on operators of this form which have the following properties:
\begin{description}
\item[Matrix-free:] One cannot easily access kernel entries $\Phi(y,x)$. Access to $\mathcal{A}$ is available only through application of $\mathcal{A}$ and $\mathcal{A}^T$ to arbitrary functions (i.e., operator actions). That is, evaluation of the maps
\begin{equation*}
	u \mapsto\mathcal{A}u \quad \text{and} \quad v \mapsto\mathcal{A}^Tv.
\end{equation*} 
\item[Non-negative kernel:] For all $(y,x) \in \Omega \times \Omega$, we have
\begin{equation*}
	\Phi(y,x) \ge 0.
\end{equation*}
\item[Local support:] Let $\phi_x$ be the following \emph{impulse response}:
\begin{equation}
\label{eq:impulse_response_defn}
\phi_x(y) := \Phi(y, x).
\end{equation} 
We say that $\mathcal{A}$ has local support if, for all $x\in \Omega$, the support of $\phi_x$ is contained (or approximately contained) in a neighborhood of $x$. The smaller these supports are, the better our algorithm will perform.
\item[Local translation invariance:] If $h$ is not too large, then
\begin{equation}
	\label{eq:local_translation_invariance}
	\Phi(y+h, x+h) \approx \Phi(y,x).
\end{equation}
\end{description}

@@

\begin{thm}[Optimal interpolation]
	\label{thm:smoothest_interpolant}
Let $y \in \mathbb{R}^d$, and let $\gamma_y$ be the function
\begin{equation}
\label{eq:u_defn_in_thm}
\gamma_y(x) := \sum_{i=1}^r \varphi_i(y)w_i(x) = \AkerPc\left(y+x,x\right),
\end{equation}
where the weighting functions $w_i$ are the solutions to optimization problem \eqref{eq:wi_optimization_problem}. Then $\gamma_y$ solves the optimization problem
\begin{equation}
\label{eq:uy_optimization_problem}
\begin{aligned}
\min_{\gamma_y \in H^2_N(\Omega)} &\quad \frac{1}{2}\norm{\Delta \gamma_y}_{L^2(\Omega)}^2 \\
\text{such that} &\quad \gamma_y(x_i) = \varphi_i(y), \quad i=1,\dots,q.
\end{aligned}
\end{equation}
\end{thm}

@@

We now describe how to efficiently evaluate $\widetilde{\Phi}(y,x)$ at an arbitrary point $(x,y) \in \Omega \times \Omega$. From \eqref{eq:kernel_entries}, we have
\begin{equation}
\label{eq:eval_y_x_000}
\AkerPc(y,x) = \sum_{i=1}^r w_i(x) \varphi_{x_i}(y-x),
\end{equation}
and from \eqref{eq:varphi_eval}, we have
\begin{equation}
\label{eq:kernel_entry_formua}
\varphi_{x_i}(y-x) = \begin{cases}
\eta^{b(i)}(y - x + x_i), & y - x + x_i \in E_{x_i} \\
0, & \text{otherwise},
\end{cases}
\end{equation}
where $b(i)$ denotes the batch index such that $x_i \in S^{b(i)}$. To compute $z=\AkerPc(y,x)$, we start with $z=0$, then loop through all batches $b$, and all sample points in each batch. For each sample point, we add one term in the sum in \eqref{eq:eval_y_x} to $z$, using \eqref{eq:kernel_entry_formua} as needed to evaluate $\varphi_{x_i}$. This is shown in Algorithm \ref{alg:eval_y_x}.

@@

\subsection{Hierarchical matrix construction}
\label{sec:H_matrix_conversion}

In this section we show how to construct an H-matrix representation of our product-convolution approximation, after discretizing the product-convolution approximation with finite elements. Let $f_1, f_2, \dots, f_N$ be a set of finite element basis functions used to discretize the problem, and let $V_h := \Span\left(f_1, f_2, \dots, f_N\right)$ be the corresponding finite element space. Here $h$ is the mesh size parameter for the finite element mesh used to discretize the problem. Further, let $p_i \in \mathbb{R}^d$, $i=1,\dots, N$ be the Lagrange nodes associated with the functions $f_i$. That is, the points such that $f_i(p_j)$ equals one if $i=j$, and zero otherwise.

In Section \ref{sec:intro} we noted that the original integral kernel, $\Aker$, is inaccessible, while in Equation \eqref{eq:kernel_entries}  we saw that the integral kernel for our product-convolution approximation, $\AkerPc \approx \Aker$, is accessible, and may be cheaply evaluated at arbitrary pairs of points $(y,x) \in \Omega \times \Omega$ via the following formula:
\begin{equation}
\label{eq:eval_y_x}
\AkerPc(y,x) = \sum_{i=1}^r w_i(x) \varphi_{x_i}(y-x).
\end{equation}
We now define a further approximation, $\AkerPcMesh \approx \AkerPc$,
\begin{equation}
\label{eq:defn_of_Akerpcmesh}
\AkerPcMesh(y,x) := \sum_{i=1}^N \sum_{j=1}^N \AkerPc(p_i,p_j) f_i(y) f_j(x)
\end{equation}
which is the interpolation of $\AkerPc$ onto $V_h \otimes V_h$. Error in the combined approximation, $\AkerPcMesh \approx \Aker$, arises both because of error in the product-convolution approximation, and because of finite element discretization error incurred from interpolation onto $V_h \otimes V_h$.

Replacing $\Aker$ with $\AkerPcMesh$ within the definition of $\Aop$ in \eqref{eq:kernel_representation} yields the following operator approximation:
\begin{equation*}
\left(\AopPcMesh u \right)(v) := \int_\Omega \int_\Omega v(y) \AkerPcMesh(y,x) u(x) dx dy.
\end{equation*}
The operator $\AopPcMesh$ maps $L^2(\Omega) \rightarrow L^2(\Omega)'$, but we are interested in the restriction of this operator to functions in the finite element space (i.e., $\AopPcMesh : V_h \rightarrow V_h'$).
%Error in the approximation $\Aop \approx \AopPcMesh$ is a combination of error in the product-convolution approximation, and error due to finite element discretization.
Let $u_h \in V_h$ and $v_h \in V_h$, and let $\mathbf{u}$ and $\mathbf{v}$ be the coefficient vectors for $u_h$ and $v_h$, respectively, with respect to the finite element basis $f_1, \dots, f_N$. I.e., $u_h(x) = \sum_{i=1}^N \mathbf{u}_i f_i(x)$, and similar for $v_h$. By using the definition of $\AkerPcMesh$ in \eqref{eq:defn_of_Akerpcmesh} and performing algebraic manipulations, it is straightforward but tedious to show that
\begin{equation*}
\left(\AopPcMesh u_h \right)(v_h) = \mathbf{v}^T \mathbf{M} \AkerPcMat \mathbf{M} \mathbf{u},
\end{equation*}
%\begin{align*}
%\left(\AopPcMesh u_h \right)(v_h) =& \int_\Omega \int_\Omega v_h(y) \AkerPcMesh(y,x) u_h(x) dx dy \\
%=& \int_\Omega \int_\Omega \left(\sum_{i=1}^N \mathbf{v}_i f_i(y)\right) \left(\sum_{j=1}^N \sum_{k=1}^N \AkerPc(p_j,p_k) f_j(y) f_k(x)\right) \left(\sum_{l=1}^N \mathbf{u}_l f_l(x)\right) dx dy \\
%=& \sum_{i,j,k,l=1}^N \mathbf{v}_i \left(\int_\Omega f_i(y) f_j(y) dy\right) \AkerPcMesh(p_j, p_k) \left(\int_\Omega f_k(x) f_l(x) dx\right) \mathbf{u}_l \\
%=& \sum_{i,j,k,l=1}^N \mathbf{v}_i \mathbf{M}_{ij} \left(\AkerPcMat\right)_{jk} \mathbf{M}_{kl} \mathbf{u}_l \\
%=& \mathbf{v}^T \mathbf{M} \AkerPcMat \mathbf{M} \mathbf{u},
%\end{align*}
where $\mathbf{M} \in \mathbb{R}^{N \times N}$ is the finite element mass matrix, which has entries
\begin{equation*}
\mathbf{M}_{ij} := \int_\Omega f_i(x) f_j(x) dx
\end{equation*}
and $\AkerPcMat \in \mathbb{R}^{N \times N}$ is the evaluation of $\AkerPcMesh$ at all pairs of finite element Lagrange nodes, which has entries 
\begin{equation*}
\left(\AkerPcMat\right)_{ij} := \AkerPcMesh(p_j, p_k).
\end{equation*}
The finite element matrix representation of $\AopPcMesh$, which we denote by $\AmatPc$, is therefore given by
\begin{equation}
\label{eq:Amatpc_defn}
\AmatPc := \mathbf{M} \AkerPcMat \mathbf{M}.
\end{equation}
We convert $\mathbf{M}$ and $\AkerPcMat$ to H-matrix format, then multiply these matrices in \eqref{eq:Amatpc_defn} using fast H-matrix methods to form an H-matrix representation of $\AmatPc$. Once in H-matrix format, useful matrix operations such as matrix factorization and inversion may be performed quickly using fast H-matrix methods. We use H1 matrices in our numerical results, but any of the other H-matrix formats (such as H2, HODLR, HSS, HBS, and others \cite{HMATRIX}) could be used instead.

The basic ideas behind H-matrices are as follows:
\begin{itemize} 
	\item Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank.
	\item Recursive algorithms can take advantage of this low rank block structure to perform matrix arithmetic fast. Conventional algorithms for matrix inversion, matrix factorization, matrix-vector products, matrix-matrix products, and matrix addition require either $O(N^2)$ or $O(N^3)$ time and memory, while the aforementioned recursive algorithms can perform these matrix operations in $O(N \log^a N)$ time and memory for hierarchical matrices. Here $N$ is the number of degrees of freedom for the discretized version of the problem, and $a=0, 1, 2$, or $3$ depending on the operation and type of H-matrix.
\end{itemize}
For more details on H-matrices, we recommend \cite{HMATRIXGOOD}.

The mass matrix, $\mathbf{M}$, is sparse, and may therefore be converted to H-matrix format using standard H-matrix methods for sparse matrices. We convert $\AkerPcMat$ into H1 matrix format using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package \cite{HLIBPRO}. Briefly, this method uses geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix (here, these locations are finite element the Lagrange nodes $p_i$, $i=1,\dots,N$) to construct a recursive block partitioning of $\AmatPc$, such that all blocks at the finest level of the partitioning are either low rank, or are small. Then, the method forms low rank approximations of the blocks that are expected to be low rank by sampling specially chosen rows and columns of those blocks. The small high rank blocks are formed directly by evaluating their entries. We describe the H-matrix construction in detail in Appendix \ref{app:h_matrix}. Although $\AkerPcMat$ is dense, constructing the H-matrix approximation via this process only requires evaluation of $O(N \log N)$ entries of $\AkerPc$, and these entries are computed via the formula in Equation \eqref{eq:eval_y_x}.

@@

In many problems of practical interest (e.g., Hessian approximation) $\Amat$ is symmetric positive semi-definite. However, $\AmatPc$ is generally non-symmetric and indefinite because of errors in the product-convolution approximation. 
This is undesirable. Symmetry and positive semi-definiteness are important properties which should be preserved if possible. Also, lacking these properties may prevent one from using highly effective algorithms, such as the conjugate gradient method, to perform further useful operations involving $\AmatPc$.
In this section we describe a method for modifying $\AmatPc$ to make it symmetric positive semi-definite.

First, we use fast H-matrix addition to symmetrize $\AmatPc$:
\begin{equation*}
\AmatPcSym := \frac{1}{2}\left(\AmatPc + \AmatPc^T\right).
\end{equation*}
Then we modify $\AmatPcSym$ to make it positive semi-definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\AmatPcSymPlus := c_0 \mathbf{I} + c_1 \AmatPcSym + c_2 \left(\AmatPcSym + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\AmatPcSym$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the moderate and large positive eigenvalues of $\AmatPcSymPlus$ approximately equal the corresponding moderate and large positive eigenvalues of $\AmatPcSym$, while the negative and small positive eigenvalues of $\AmatPcSym$ are modified so that $\AmatPcSymPlus$ is positive semi-definite. Later in this section we will explain how $c_0$, $c_1$, $c_2$, and $\mu$ are chosen and how our choice of these scalars affects the spectral properties of $\AmatPcSymPlus$. In Figure FIG we compare the spectrum of $\AmatPcSym$ to that of $\AmatPcSymPlus$ for different choices of these scalars. Forming $\AmatPcSymPlus$ requires computing an H-matrix inverse, $\left(\AmatPcSym + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should usually be avoided, here it is computationally acceptable because $\AmatPcSym + \mu \mathbf{I}$ is an H-matrix. Inversion of H-matrices is a routine procedure which only requires $O(N k^2 \log(N)^2)$ work, where $k$ is the H-matrix rank \cite{Hmatrixinverse}.

Approximation \eqref{eq:rational_matrix_function} may be written as $\AmatPcSymPlus = f(\AmatPcSym)$, where
\begin{equation}
\label{eq:rational_function_scalar}
f(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\mu}.
\end{equation}
We choose $c_0, c_1, c_2$, and $\mu$, to be the solution to the following system of equations:
\begin{subequations}
	\label{eq:four_eq_system}
\begin{align}
f(0) &= 0 \label{eq:desired_property_1} \\
f'(0) &= 0 \label{eq:desired_property_2} \\
f(\lambda_\text{min}) &= |\lambda_\text{min}| \label{eq:desired_property_3} \\
f(\lambda_\text{max}) &= \lambda_\text{max}, \label{eq:desired_property_4}
\end{align}
\end{subequations}
where $\lambda_\text{min}$ and $\lambda_\text{max}$ are the smallest and largest eigenvalues of $\AmatPcSym$, respectively. We also require $\mu > |\lambda_\text{min}|$, so that the pole of $f$ is not in $[\lambda_\text{min},\lambda_\text{max}]$. Note that $\lambda_\text{min} < 0$ since $\AmatPcSym$ is indefinite. 
We estimate $\lambda_\text{min}$ and $\lambda_\text{max}$ using the implicitly restarted Lanczos method \cite{lanczos,scipy}.

Together, equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, \eqref{eq:desired_property_3},  and \eqref{eq:desired_property_4}, and the condition $\mu > |\lambda_\text{min}|$, imply that $f(0)=0$ is the unique minimum of $f$ on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that $f$ is non-negative on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that  $\AmatPcSymPlus$ is positive semi-definite. Furthermore, \eqref{eq:desired_property_1} implies that the null space of $\AmatPcSymPlus$ contains the null-space of $\AmatPcSym$. This is important when the operator $\mathcal{A}$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. Equation \eqref{eq:desired_property_4} forces the moderate and large positive eigenvalues of $\AmatPcSymPlus$ to be close to the corresponding moderate and large positive eigenvalues of $\AmatPcSym$. I.e., $f$ modifies the ``important'' part of the spectrum of $\AmatPcSym$ as little as possible. Equation \eqref{eq:desired_property_3} ensures that the negative eigenvalues of $\AmatPcSym$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $f$. There is a tradeoff: if we choose a larger value for $f(\lambda_\text{min})$, then the positive eigenvalues of $\AmatPcSymPlus$ will be closer to the corresponding positive eigenvalues of $\AmatPcSym$ (good), but the erroneous negative eigenvalues of $\AmatPcSym$ will be amplified in magnitude (bad). This is shown in Figure FIG. We find that $f(\lambda_\text{min})=|\lambda_\text{min}|$ is an appropriate balance between these competing interests.

Solving system \eqref{eq:four_eq_system} is computationally trivial. Using the definition of $f$ in \eqref{eq:rational_function_scalar}, and basic calculus and linear algebra, we may write Equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, and \eqref{eq:desired_property_4} in the following matrix form:
\begin{equation}
\label{eq:rational_3_linear_system}
\begin{bmatrix} 1 & 0 & 1/\mu \\ 0 & 1 & -1/\mu^2 \\ 1 & \lambda_\text{max} & 1/(\lambda_\text{max}+\mu) \end{bmatrix} \begin{bmatrix} c_0 \\ c_1 \\ c_2 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \\ \lambda_\text{max} \end{bmatrix},
\end{equation}
and we may write Equation \eqref{eq:desired_property_3} as follows: 
\begin{equation}
\label{eq:rational_extra_eq}
c_0 + c_1 \lambda_\text{min} + \frac{c_2}{\lambda_\text{min}+\mu} - |\lambda_\text{min}| = 0.
\end{equation}
Let 
\begin{equation*}
g(\mu) := c_0(\mu) + c_1(\mu) \lambda_\text{min} + \frac{c_2(\mu)}{\lambda_\text{min}+\mu} - |\lambda_\text{min}|
\end{equation*}
denote the left hand side of \eqref{eq:rational_extra_eq}, but with $c_0(\mu)$, $c_1(\mu)$, and $c_2(\mu)$ being the solution of \eqref{eq:rational_3_linear_system} for the given value of $\mu$. I.e., within $g(\mu)$ the quantities $c_0$, $c_1$, and $c_2$ are no longer free variables; they now depend on $\mu$ implicitly through the solution of a linear system. We solve the combined system, \eqref{eq:rational_3_linear_system} and $\eqref{eq:rational_extra_eq}$, by solving the one-dimensional nonlinear root finding problem
\begin{equation}
\label{eq:g_mu_zero}
g(\mu)=0
\end{equation}
with bound $\mu > |\lambda_\text{min}|$ and initial guess $\mu_0 = 2 |\lambda_\text{min}|$. Each evaluation of $g$ requires solving the $3 \times 3$ linear system \eqref{eq:rational_3_linear_system}, and once the solution $\mu$ is found, we solve \eqref{eq:rational_3_linear_system} one more time to get the final values of $c_0$, $c_1$, and $c_2$.
